Simulating High-Dimensional Multivariate Data using the \pkg{bigsimr} Package

### 2021-05-25 article_bigsimr.tex
Revise for JSS submission up to Chapter 4

\author{
Alfred G. Schissler\\University of Nevada \And Edward J.
Bedrick\\University of Nevada \And Alexander D. Knudson\\University of
Nevada \AND Tomasz J. Kozubowski\\University of Nevada \And Tin
Nguyen\\University of Nevada \And Anna K. Panorska\\University of Nevada
\AND Juli Petereit\\University of Nevada \And Walter W.
Piegorsch\\University of Nevada \And Duc Tran\\University of Nevada
}
\title{Simulating High-Dimensional Multivariate Data using the
\pkg{Bigsimr} Package}

\Plainauthor{Alfred G. Schissler, Edward J. Bedrick, Alexander D.
Knudson, Tomasz J. Kozubowski, Tin Nguyen, Anna K. Panorska, Juli
Petereit, Walter W. Piegorsch, Duc Tran}
\Plaintitle{Simulating High-Dimensional Multivariate Data using the
Bigsimr Package}
\Shorttitle{\pkg{Bigsimr}: Simulate High-Dimensional Data}

\Abstract{
It is critical to realistically simulate data when employing Monte Carlo
techniques and evaluating statistical methodology. Measurements are
often correlated and high dimensional in this era of big data, such as
data obtained through high-throughput biomedical experiments. Due to
computational complexity and a lack of user-friendly software available
to simulate these massive multivariate constructions, researchers resort
to simulation designs that posit independence or perform arbitrary data
transformations. This article introduces the
\textbackslash pkg\{Bigsimr\} R package. This high-level package
provides a flexible and scalable procedure to simulate high-dimensional
random vectors with arbitrary marginal distributions and known Pearson,
Spearman, or Kendall correlation matrix. \textbackslash pkg\{Bigsimr\}
contains additional high-performance features, including multi-core
algorithms to simulate random vectors, estimate correlation, and compute
the nearest correlation matrix. Monte Carlo studies quantify the
accuracy and scalability of our approach, up to \(d=10,000\). Finally,
we demonstrate example applications enabled via
\textbackslash pkg\{Bigsimr\} by applying these methods to a motivating
dataset --- RNA-sequencing data obtained from breast cancer tumor
samples.
}

@@

Massive high-dimensional (HD) data sets are now commonplace in many
areas of scientific inquiry. As new methods are developed for data, a
fundamental challenge lies in designing and conducting simulation
studies to assess the operating characteristics of analyzing such
proposed methodology --- such as false positive rates, statistical
power, interval coverage, and robustness --- often with comparison to
existing methods. Further, efficient simulation empowers statistical
computing strategies, such as the parametric bootstrap
\citep{Chernick2008} to simulate from a hypothesized null model,
providing inference in analytically challenging settings. Such Monte
Carlo (MC) techniques become difficult for high-dimensional data using
existing algorithms and tools. This is particularly true when simulating
massive multivariate, non-normal distributions, arising naturally in
many fields of study.

As many have noted, it can be vexing to simulate dependent,
non-normal/discrete data, even for low dimensional settings
\citep{MB13, XZ19}. For continuous non-normal multivariate data, the
well-known NORmal To Anything (NORTA) algorithm \citep{Cario1997} and
other copula approaches \citep{Nelsen2007} are well-studied, with
flexible, robust software available \citep{Yan2007, Chen2001}. Yet these
approaches do not scale in a timely fashion to high-dimensional problems
\citep{Li2019gpu}. For discrete data, early simulation strategies had
major flaws, such as failing to obtain the full range of possible
correlations (e.g., admitting only positive correlations: see
\citet{Park1996}). While more recent approaches
\citep{MB13, Xia17, BF17} have largely remedied this issue for
low-dimensional problems, the existing tools are not designed to scale
to high dimensions.

Another central issue lies in characterizing dependence between
components in the high-dimensional random vector. The choice of
correlation in practice usually relates to the eventual analytic goal
and distributional assumptions of the data (e.g., non-normal, discrete,
infinite support, etc). For normal data, the Pearson product-moment
correlation describes the dependency perfectly. As we will see, however,
simulating arbitrary random vectors that match a target Pearson
correlation matrix is computationally intense \citep{Chen2001, Xia17}.
On the other hand, an analyst might consider use of nonparametric
correlation measures to better characterize monotone, non-linear
dependence, such as Spearman's \(\rho\) and Kendall's \(\tau\).
Throughout, we focus on matching these nonparametric dependence
measures, as our aim lies in modeling non-normal data and these
rank-based measures possess invariance properties favorable in our
proposed methodology. We do, however, provide Pearson matching with some
caveats.

With all this in mind, we present a scalable, flexible multivariate
simulation algorithm. The crux of the method lies in the construction of
a Gaussian copula in the spirit of the NORTA procedure. Further, we
introduce the \texttt{bigsimr} R package that provides high-performance
software implementing our algorithm. The algorithm design leverages
useful properties of nonparametric correlation measures, namely
invariance under monotone transformation and well-known closed-form
relationships between dependence measures for the multivariate normal
(MVN) distribution.

Our study proceeds by providing background information, including a
description of a motivating example application: RNA-sequencing
(RNA-seq) breast cancer data. Then we describe and justify our
simulation methodology and related algorithms. Next, we detail an
illustrative low-dimensional example of basic use of the
\texttt{bigsimr} R package. Then we proceed with Monte Carlo studies
under various bivariate distributional assumptions to assess accuracy.
We conclude the Monte Carlo evaluations by summarizing the computation
time for increasingly higher dimensional vectors. After the MC
evaluations, we simulate random vectors motivated by our RNA-seq
example, evaluate the accuracy, and provide example statistical
computing tasks, namely MC estimation of joint probabilities and
evaluating HD correlation estimation efficiency. Finally, we discuss the
method's utility, limitations, and future directions.

@@

The \texttt{bigsimr} \texttt{R} package presented here provides multiple
algorithms that operate with high-dimensional multivariate data;
however, all these algorithms were originally designed to support a
single task: to generate random vectors drawn from multivariate
probability distributions with given marginal distributions and
dependency metrics. Specifically, our goal is to efficiently simulate a
large number, \(B\), of HD random vectors
\({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with \emph{correlated} components
and heterogeneous marginal distributions, described via cumulative
distribution functions (CDFs) \(F_i\).

@@

\begin{itemize}
\tightlist
\item
  BP1: A wide range of dependences, allowing both positive and negative
  values, and, ideally, admitting the full range of possible values.
\item
  BP2: Flexible dependence, meaning that the number of bivariate
  marginals can be equal to the number of dependence parameters.
\item
  BP3: Flexible marginal modeling, generating heterogeneous data ---
  possibly from differing probability families.
\end{itemize}

@@

Simulating high-dimensional, non-normal, correlated data motivates this
work --- in pursuit of modeling RNA-sequencing (RNA-seq) data
\citep{Wang2009b, Conesa2016b} derived from breast cancer patients. The
RNA-seq data-generating process involves counting how often a particular
form of human messenger RNA (mRNA) is expressed in a biological sample.
RNA-seq platforms typically quantify the entire transcriptome in one
experimental run, resulting in high-dimensional data. For human-derived
samples, this results in count data corresponding to over 20,000 genes
(protein-coding genomic regions) or even over 77,000 isoforms when
alternatively-spliced mRNA are counted \citep{Schissler2019}.
Importantly, due to inherent biological processes, gene expression data
exhibit correlation (co-expression) across genes
\citep{BE07, Schissler2018}.

We illustrate our methodology using a well-studied Breast Invasive
Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see
Acknowledgments). For ease of modeling and simplicity of exposition, we
only consider high expressing genes. In turn, we begin by filtering to
retain the top 1000 of the highest-expressing genes (in terms of median
expression) of the over 20,000 gene measurements from \(N=878\)
patients' tumor samples. This gives a great number of pairwise
dependencies among the marginals (specifically,
\(\ensuremath{4.995\times 10^{5}}\) correlation parameters). Table
@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected
high-expressing genes for the first five patients' breast tumor samples.
To help visualize the bivariate relationships for these three selected
genes across all patients, Figure @ref(fig:ch010-realDataFig) displays
the marginal distributions and estimated Spearman's correlations.

@@

\caption{\label{tab:ch010-realDataTab}mRNA expression for three selected high-expressing genes, STAU1, FKBP1A, NME2, for the first five patients in the TCGA BRCA data set.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
Patient ID & STAU1 & FKBP1A & NME2\\
\midrule
TCGA-A1-A0SB & 10440 & 11354 & 17655\\
TCGA-A1-A0SD & 21523 & 20221 & 14653\\
TCGA-A1-A0SE & 21733 & 22937 & 35251\\
TCGA-A1-A0SF & 11866 & 19650 & 16551\\
TCGA-A1-A0SG & 12486 & 12089 & 10434\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption[Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1]{Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes from Table 1. The data possess outliers, heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology.}\label{fig:ch010-realDataFig}
\end{figure}
\end{CodeChunk}

@@

In multivariate analysis, an analyst must select a metric to quantify
dependency. The most widely-known is the Pearson (product-moment)
correlation coefficient that describes the linear association between
two random variables \(X\) and \(Y\), and, it is given by

@@

As \citet{MB13} and \citet{MK01} discuss, for a bivariate normal
\((X,Y)\) random vector, the Pearson correlation completely describes
the dependency between the components. For non-normal marginals with
monotone correlation patterns, however, \(\rho_P\) suffers some
drawbacks and may mislead or fail to capture important relationships
\citep{MK01}. Alternatively in these settings, analysts often prefer
rank-based correlation measures to describe the degree of monotonic
association.

Two nonparametric, rank-based measures common in practice are Spearman's
correlation (denoted \(\rho_S\)) and Kendall's \(\tau\). Spearman's
\(\rho_S\) has an appealing correspondence as the Pearson correlation
coefficient on \emph{ranks} of the values, thereby captures nonlinear
yet monotone relationships. Kendall's \(\tau\), on the other hand, is
the difference in probabilities of concordant and discordant pairs of
observations, \((X_i, Y_i)\) and \((X_j, Y_j)\), with concordance
meaning that orderings have the same direction (e.g., if \(X_i < X_j\),
then \(Y_i < Y_j\)). Note that concordance is determined by the ranks of
the values, not the values themselves.

@@

Given two marginal distributions, \(\rho_P\) is not free to vary over
the entire range of possible correlations \([-1,1]\). The so-called
\emph{Frechet-Hoeffding bounds} are well-studied
\citep{Nelsen2007, BF17}. These constraints cannot be overcome through
algorithm design. In general, the bounds are given by

@@

There is a strong connection of our simulation strategy to Gaussian
\emph{copulas} (see \citet{Nelsen2007} for a technical introduction). A
copula is a distribution function on \([0,1]^d\) that describes a
multivariate probability distribution with standard uniform marginals.
This provides a powerful, natural way to characterize joint probability
structures. Consequently, the study of copulas is an important and
active area of statistical theory and practice.

@@

A Gaussian copula has marginal CDFs that are all standard normal,
\(F_i = \Phi, \forall \, i\). This representation corresponds to a
multivariate normal (MVN) distribution with standard normal marginal
distributions and covariance matrix \({\bf R_P}\). As the marginals are
standardized to have unit variance, however, \({\bf R_P}\) is a Pearson
correlation matrix. If \(F_{{\bf R}}\) is the CDF of such a multivariate
normal distribution, then the corresponding Gaussian copula
\(C_{{\bf R}}\) is defined through

@@

where \(\Phi(\cdot)\) is the standard normal CDF. Note that the copula
\(C_{{\bf R}}\) is the familiar multivariate normal CDF of the random
vector \((\Phi(X_1), \ldots, \Phi(X_d))\), where
\((X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})\).

@@

\hypertarget{relevant-multivariate-simulation-software-comparison}{%
\subsection{Relevant Multivariate Simulation software
Comparison}\label{relevant-multivariate-simulation-software-comparison}}

To place \texttt{Bigsimr}'s implementation and algorithms in context
with existing approaches, as well as make this work's contribution more
evident, we make some qualitative comparisons to other \texttt{R}
multivariate simulation packages. The \texttt{MASS} packages
\texttt{mvnorm()} has long been available to produce random vectors from
multivariate normal (and \(t\)) distributions. This familiar procedure
cannot simulate from arbitrary margins and dependency measures, as well
as fails to readily scale to high dimensions. Recently, the
high-performance package \texttt{mvnfast} has been released that
provides scaleable multivariate normal generation, but this doesn't
completely satistify our goals of flexible margins and dependency
modeling. For dependent discrete data, small-dimesional packages exist,
such as \texttt{GenOrd}. The full-featured package \texttt{copula}
provides many different copulas and a fairly general random generator
\texttt{rCopula}. For low-dimensional modeling situtations, copula is
suitable, yet the entent of exact matching of margins automatically is
unclear. High-dimesional random vectors via \texttt{copula} takes an
impractical amount of computing time \citep{Li2019gpu}. The
\texttt{nortaRA} R package matches Pearson coefficients near exactly and
is reasonable for low-dimenional settings. In sum, we know of no other
\texttt{R} package that can produce HD random vectors with flexible
margins and dependence modeling. Table (\#tab:compare-table) summarizes
the above discussion.

\begin{table}[]
\centering
\caption{Selected feature comparison of `bigsimr` to other existing `R` multivariate simulation packages.}
\label{tab:compare-table}
\begin{tabular}{@{}llll@{}}
R Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
MASS      & No           & No                    & No                  \\
mvnfast   & Yes          & No                    & No                  \\
GenOrd    & No           & Discrete margins      & Yes                 \\
copula    & No           & Yes                   & No                  \\
nortaRA   & No           & Yes                   & No                  \\
bigsimr   & Yes          & Yes                   & Yes                
\end{tabular}
\end{table}

@@

This section describes our methods involved in simulating a random
vector \(\bf Y\) with \(Y_i\) components for \(i=1,\ldots,d\). Each
\(Y_i\) has a specified marginal CDF \(F_i\) and its inverse
\(F^{-1}_i\). To characterize dependency, every pair \((Y_i, Y_j)\) has
a given Pearson correlation \(\rho_P\), Spearman correlation \(\rho_S\),
and/or Kendall's \(\tau\). The method can be described as a
\emph{high-performance Gaussian copula} (Equation @ref(eq:gauss))
providing a high-dimensional NORTA-inspired algorithm.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulate a random vector \(\bf Z\) with \(d\) \emph{independent and
  identically distributed} (iid) standard normal components.
\item
  Determine the input matrix \(\Sigma_{\bf Z}\) that corresponds with
  the specified output \(\Sigma_{\bf Y}\).
\item
  Produce a Cholesky factor \(M\) of \(\Sigma_{\bf Z}\) such that
  \(M M^{\prime}=\Sigma_{\bf Z}\).
\item
  Set \(X\) by \(X \gets MZ\).
\item
  \(\text{Return} \; Y \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d\).
\end{enumerate}

@@

With modern parallel computing, steps 1, 3, 4, 5 are readily implemented
as high-performance, multi-core and/or graphical-processing-unit (GPU)
accelerated algorithms --- providing fast scalability using
readily-available hardware.

Matching specified Pearson correlation coefficients exactly (step 2
above), however, is computationally costly. In general, there is no
closed-form correspondence between the components of the input
\(\Sigma_{\bf Z}\) and target \(\Sigma_{\bf Y}\). Matching the
correlations involves evaluating or approximating \(\binom{d}{2}\)
integrals of the form

@@

where \(\phi(\cdot)\) is the joint probability distribution function of
two correlated standard normal variables. For HD data, (nearly) exact
evaluation may become too costly to enable practical simulation studies.
For low-dimensional problems, however, methods and tools exist to match
Pearson correlations precisely; see \citep{Xia17} and the publicly
available \texttt{nortaRA} R package \citep{Chen2001}. As described
later, to enable HD Pearson matching we approximate these integrals.

@@

Sklar's theorem provides a useful characterization of multivariate
distributions through copulas. Yet the choice of copula-based simulation
algorithm affects which joint distributions may be simulated. Even in
low-dimensional spaces (e.g., \(d=3\)), there exist valid multivariate
distributions with \emph{feasible} Pearson correlation matrices that
NORTA cannot match exactly \citep{LH75}. This occurs when the bivariate
transformations are applied to find the input correlation matrix, yet
when combined the resultant matrix is indefinite. These situations do
occur, even using exact analytic calculations. Such problematic target
correlation matrices are termed \emph{NORTA defective}.

@@

There is hope, however, as \citet{GH02} also showed that replacing an
indefinite input correlation matrix with a close proxy will give
approximate matching to the target --- with adequate performance for
moderate \(d\). This provides evidence that our nearest positive
definite (PD) augmented approach will maintain reasonable accuracy if
our input matching scheme returns an indefinite matrix, at least for the
rank-based matching scheme described above.

\hypertarget{rand-vec-gen}{%
\subsection{Random vector generator}\label{rand-vec-gen}}

We now describe our algorithm to generate random vectors, which mirrors
the classical NORTA algorithm above with some modifications for
rank-based dependency matching:

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Mapping step

>>>>

\author{
Alfred G. Schissler\\University of Nevada \And Edward J.
Bedrick\\University of Arizona \And Alexander D. Knudson\\University of
Nevada \AND Tomasz J. Kozubowski\\University of Nevada \And Tin
Nguyen\\University of Nevada \And Anna K. Panorska\\University of Nevada
\AND Juli Petereit\\University of Nevada \And Walter W.
Piegorsch\\University of Arizona \And Duc Tran\\University of Nevada
}
\title{Simulating High-Dimensional Multivariate Data using the
\pkg{bigsimr} Package}

\Plainauthor{Alfred G. Schissler, Edward J. Bedrick, Alexander D.
Knudson, Tomasz J. Kozubowski, Tin Nguyen, Anna K. Panorska, Juli
Petereit, Walter W. Piegorsch, Duc Tran}
\Plaintitle{Simulating High-Dimensional Multivariate Data using the
bigsimr Package}
\Shorttitle{\pkg{bigsimr}: Simulate High-Dimensional Data}

\Abstract{
It is critical to accurately simulate data when employing Monte Carlo
techniques and evaluating statistical methodology. Measurements are
often correlated and high dimensional in this era of big data, such as
data obtained through high-throughput biomedical experiments. Due to
computational complexity and a lack of user-friendly software available
to simulate these massive multivariate constructions, researchers resort
to simulation designs that posit independence or perform arbitrary data
transformations. To meet this gap, we developed the bigsimr R package
for high-dimensional random vector simulation with arbitrary marginal
distributions and known Pearson, Spearman, or Kendall correlation
matrix. bigsimr contains additional high-performance features, including
multi-core algorithms to estimate correlation and compute the nearest
correlation matrix. Monte Carlo studies quantify the accuracy and
scalability of our approach, up to \(d=10,000\). Finally, we describe
example workflows and apply to a motivating high-dimensional data set
--- RNA-sequencing data obtained from breast cancer tumor samples.
}

@@

Massive high-dimensional (HD) data sets are now commonplace in many
areas of scientific inquiry. As new methods are developed for data
analysis, a fundamental challenge lies in designing and conducting
simulation studies to assess the operating characteristics of proposed
methodology --- such as false positive rates, statistical power,
interval coverage, and robustness --- often to compare to methods.
Further, efficient simulation empowers statistical computing strategies,
such as the parametric bootstrap \citep{Chernick2008} to simulate from a
hypothesized null model, providing inference in analytically challenging
settings. Such Monte Carlo (MC) techniques become difficult for HD
dependent data using existing algorithms and tools. This is particularly
true when simulating massive multivariate, non-normal distributions,
arising in many fields of study.

As many have noted, it can be vexing to simulate dependent,
non-normal/discrete data, even for low-dimensional (LD) settings
\citep{MB13, XZ19}. For continuous non-normal LD multivariate data, the
well-known NORmal To Anything (NORTA) algorithm \citep{Cario1997} and
other copula approaches \citep{Nelsen2007} are well-studied, with
flexible, robust software available \citep{Yan2007, Chen2001}. Yet these
approaches do not scale in a timely fashion to HD problems
\citep{Li2019gpu}. For discrete data, early simulation strategies had
major flaws, such as failing to obtain the full range of possible
correlations (e.g., admitting only positive correlations: see
\citet{Park1996}). While more recent approaches
\citep{MB13, Xia17, BF17} have largely remedied this issue for LD
problems, the existing tools are not designed to scale to high
dimensions for discrete marginal distributions.

Another central issue lies in characterizing dependence between
components in the HD random vector. The choice of correlation in
practice usually relates to the eventual analytic goal and
distributional assumptions of the data (e.g., non-normal, discrete,
infinite support, etc). For normal data, the Pearson product-moment
correlation describes the dependency perfectly. As we will see, however,
simulating arbitrary random vectors that match a target Pearson
correlation matrix is computationally intense \citep{Chen2001, Xia17}.
On the other hand, an analyst might consider use of nonparametric
correlation measures to better characterize monotone, non-linear
dependence, such as Spearman's \(\rho\) and Kendall's \(\tau\).
Throughout, we focus on matching these nonparametric dependence
measures, as our aim lies in modeling non-normal data and these
rank-based measures possess invariance properties favorable in our
proposed methodology. We do, however, implement Pearson matching, but
several layers of approximation is required.

With all this in mind, we present a scalable, flexible multivariate
simulation algorithm. The crux of the method lies in the construction of
a Gaussian copula in the spirit of the NORTA procedure. As we will
describe in more detail, the algorithm's design leverages useful
properties of nonparametric correlation measures, namely invariance
under monotone transformation and well-known closed-form relationships
between dependence measures for the multivariate normal (MVN)
distribution. For our method, we developed a high-performance
implementation: the \texttt{bigsimr} R package, written in native Julia
code.

Our study proceeds by providing background information, including a
description of a motivating example application: RNA-sequencing
(RNA-seq) breast cancer data. Then we describe and justify our
simulation methodology and related algorithms. We proceed by providing
an illustrative LD \texttt{bigsimr} workflow. Next we conduct MC studies
under various bivariate distributional assumptions to evaluate
performance and accuracy. After the MC evaluations, we simulate random
vectors motivated by our RNA-seq example, evaluate the accuracy, and
provide example statistical computing tasks, namely MC estimation of
joint probabilities and evaluating HD correlation estimation efficiency.
Finally, we discuss the method's utility, limitations, and future
directions.

@@

The \texttt{bigsimr} \texttt{R} package presented here provides multiple
algorithms that operate with HD multivariate data; however, all these
algorithms were originally designed to support a single task: to
generate random vectors drawn from multivariate probability
distributions with given marginal distributions and dependency metrics.
Specifically, our goal is to efficiently simulate a large number, \(B\),
of HD random vectors \({\bf Y}=(Y_1, \ldots, Y_d)^\top\) with
\emph{correlated} components and heterogeneous marginal distributions,
described via cumulative distribution functions (CDFs)
\(F_i, i=1,\ldots,d\).

@@

\begin{itemize}
\tightlist
\item
  BP1: A wide range of dependences, allowing both positive and negative
  values, and, ideally, admitting the full range of possible values.
\item
  BP2: Flexible dependence, meaning that the number of bivariate
  marginals can be equal to the number of dependence parameters.
\item
  BP3: Flexible marginal modeling, generating heterogeneous data ---
  including mixed continuous and discrete margins.
\end{itemize}

@@

Simulating HD, non-normal, correlated data motivates this work, in
pursuit of modeling RNA-sequencing (RNA-seq) data
\citep{Wang2009b, Conesa2016b} derived from breast cancer patients. The
RNA-seq data-generating process involves counting how often a particular
form of messenger RNA (mRNA) is expressed in a biological sample.
RNA-seq platforms typically quantify the entire transcriptome in one
experimental run, resulting in HD data. For human-derived samples, this
results in count data corresponding to over 20,000 genes (protein-coding
genomic regions) or even over 77,000 isoforms when alternatively-spliced
mRNA are counted \citep{Schissler2019}. Importantly, due to inherent
biological processes, gene expression data exhibit correlation
(co-expression) across genes \citep{BE07, Schissler2018}.

We illustrate our methodology using the Breast Invasive Carcinoma (BRCA)
data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
For ease of modeling and simplicity of exposition, we only consider high
expressing genes. In turn, we begin by filtering to retain the top 1000
of the highest-expressing genes (in terms of median expression) of the
over 20,000 gene measurements from \(N=878\) patients' tumor samples.
This gives a great number of pairwise dependencies among the marginals
(specifically, \(\ensuremath{4.995\times 10^{5}}\) correlation
parameters). Table @ref(tab:ch010-realDataTab) displays RNA-seq counts
for three selected high-expressing genes for the first five patients'
breast tumor samples. To help visualize the bivariate relationships for
these three selected genes across all patients, Figure
@ref(fig:ch010-realDataFig) displays the marginal distributions and
estimated Spearman's correlations.

@@

\caption{\label{tab:tab:ch010-realDataTab}mRNA expression for three selected high-expressing genes, STAU1, FKBP1A, NME2, for the first five patients in the TCGA BRCA data set.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
Patient ID & STAU1 & FKBP1A & NME2\\
\midrule
TCGA-A1-A0SB & 10440 & 11354 & 17655\\
TCGA-A1-A0SD & 21523 & 20221 & 14653\\
TCGA-A1-A0SE & 21733 & 22937 & 35251\\
TCGA-A1-A0SF & 11866 & 19650 & 16551\\
TCGA-A1-A0SG & 12486 & 12089 & 10434\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption[Pairwise scatterplots for three example genes (Table 1), with estimated Spearman's correlations and marginal density estimates displayed]{Pairwise scatterplots for three example genes (Table 1), with estimated Spearman's correlations and marginal density estimates displayed. The data possess outliers, heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology.}\label{fig:ch010-realDataFig}
\end{figure}
\end{CodeChunk}

@@

In multivariate analysis, an analyst must select a metric to quantify
dependency. The most widely-known is the Pearson correlation coefficient
that describes the linear association between two random variables \(X\)
and \(Y\), and, it is given by

@@

As \citet{MB13} and \citet{MK01} discuss, for a bivariate normal
\((X,Y)\) random vector, the Pearson correlation completely describes
the dependency between the components. For non-normal marginals with
monotone correlation patterns, however, \(\rho_P\) suffers some
drawbacks and may mislead or fail to capture important relationships
\citep{MK01}. Alternatively, analysts often prefer rank-based
correlation measures to describe the degree of monotonic association.

Two nonparametric, rank-based measures common in practice are Spearman's
correlation (denoted \(\rho_S\)) and Kendall's \(\tau\). Spearman's
\(\rho_S\) has an appealing correspondence as the Pearson correlation
coefficient on \emph{ranks} of the values, thereby capturing nonlinear
yet monotone relationships. Kendall's \(\tau\), on the other hand, is
the difference in probabilities of concordant and discordant pairs of
observations, \((X_i, Y_i)\) and \((X_j, Y_j)\) (concordance meaning
that orderings have the same direction, e.g., if \(X_i < X_j\), then
\(Y_i < Y_j\)). Note that concordance is determined by the ranks of the
values, not the values themselves.

@@

Given two marginal distributions, \(\rho_P\) is not free to vary over
the entire range of possible correlations \([-1,1]\). The
\emph{Frechet-Hoeffding bounds} are well-studied
\citep{Nelsen2007, BF17}. These constraints cannot be overcome through
algorithm design. In general, the bounds are given by

@@

There is a connection of our simulation strategy to Gaussian
\emph{copulas} \citep{Nelsen2007}. A copula is a distribution function
on \([0,1]^d\) that describes a multivariate probability distribution
with standard uniform marginals. This provides a powerful, natural way
to characterize joint probability. Consequently, the study of copulas is
an important and active area of statistical theory and practice.

@@

A Gaussian copula has marginal CDFs that are all standard normal,
\(F_i = \Phi, \forall \, i\), where \(\Phi(\cdot)\) is the standard
normal CDF. This representation corresponds to a multivariate normal
(MVN) distribution with standard normal marginal distributions and
covariance matrix \({\bf R_P}\). Since the marginals are standardized to
have unit variance, however, \({\bf R_P}\) is a Pearson correlation
matrix. If \(F_{{\bf R}}\) is the CDF of such a multivariate normal
distribution, then the corresponding Gaussian copula \(C_{{\bf R}}\) is
defined through

@@

Here the copula function \(C_{{\bf R}}\) is the familiar multivariate
normal CDF of the random vector \((\Phi(X_1), \ldots, \Phi(X_d))\),
where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})\).

@@

\hypertarget{other-multivariate-simulation-r-packages}{%
\subsection{Other Multivariate Simulation R
Packages}\label{other-multivariate-simulation-r-packages}}

To place \texttt{bigsimr}'s implementation and algorithms in context
with existing approaches, here we qualitatively compare to other
\texttt{R} multivariate simulation packages. The \texttt{MASS} packages
\texttt{mvnorm()} has long been available to produce random vectors from
multivariate normal (and \(t\)) distributions. This familiar procedure
cannot simulate from arbitrary margins and dependency measures, as well
as fails to readily scale to high dimensions. Recently, the
high-performance package \texttt{mvnfast} \citep{Fasiolo2016} has been
released that provides HD multivariate normal generation, but this lacks
flexibility in marginal models and dependency modeling. For dependent
discrete data, LD packages exist, such as \texttt{GenOrd}. For copula
computation, the full-featured package \texttt{copula} provides many
different copulas and a fairly general LD random generator
\texttt{rCopula()}. High-dimesional random vectors via \texttt{copula}
takes an impractical amount of computing time \citep{Li2019gpu}. The
\texttt{nortaRA} R package matches Pearson coefficients near exactly and
is reasonable for LD settings. Finally, we know of no other \texttt{R}
package that can produce HD random vectors with flexible margins and
dependence modeling. Table (\#tab:compare-table) summarizes the above
discussion.

\begin{table}[]
\centering
\caption{Selected feature comparison of bigsimr to existing R packages for multivariate simulation.}
\label{tab:compare-table}
\begin{tabular}{@{}llll@{}}
R Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
MASS      & No           & No                    & No                  \\
mvnfast   & Yes          & No                    & No                  \\
GenOrd    & No           & Discrete margins      & Yes                 \\
copula    & No           & Yes                   & No                  \\
nortaRA   & No           & Yes                   & No                  \\
bigsimr   & Yes          & Yes                   & Yes                
\end{tabular}
\end{table}

@@

This section describes our methods involved in simulating a random
vector \(\bf Y\) with \(Y_i\) components for \(i=1,\ldots,d\). Each
\(Y_i\) has a specified marginal CDF \(F_i\) and its inverse
\(F^{-1}_i\). To characterize dependency, every pair \((Y_i, Y_j)\) has
a given Pearson correlation \(\rho_P\), Spearman correlation \(\rho_S\),
and/or Kendall's \(\tau\). The method can be described as a
\emph{high-performance Gaussian copula} (Equation @ref(eq:gauss))
providing a HD NORTA-inspired algorithm.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulate a random vector \(\bf Z\) with \(d\) independent and
  identically distributed (iid) standard normal components.
\item
  Determine the input matrix \(\Sigma_{\bf Z}\) that corresponds with
  the specified output \(\Sigma_{\bf Y}\).
\item
  Produce a Cholesky factor \(M\) of \(\Sigma_{\bf Z}\) such that
  \(M M^{\prime}=\Sigma_{\bf Z}\).
\item
  Set \(X\) by \(X \gets MZ\).
\item
  \(\text{Return} \; Y \; \text{where} \; Y_i \gets F_i^{-1}[\Phi(X_i)], \; i=1,...,d\).
\end{enumerate}

@@

With modern parallel computing, steps 1, 3, 4, 5 are readily implemented
as high-performance, multi-core and/or graphical-processing-unit (GPU)
accelerated algorithms --- providing fast scalability to HD.

Matching specified Pearson correlation coefficients exactly (step 2
above), however, is computationally costly. In general, there is no
convenient correspondence between the components of the input
\(\Sigma_{\bf Z}\) and target \(\Sigma_{\bf Y}\). Matching the
correlations involves evaluating or approximating \(\binom{d}{2}\)
integrals of the form

@@

where \(\phi(\cdot)\) is the joint probability distribution function of
two correlated standard normal variables. For HD simulation, exact
evaluation may become too costly to enable practical simulation studies.
Our remedy to enable HD Pearson matching in \texttt{bigsimr} is Hermite
polynomial approximation of these \(\choose{d}{2}\) double integrals, as
suggested by \citet{XZ19}.

@@

Sklar's theorem provides a useful characterization of multivariate
distributions through copulas. Yet the choice of copula-based simulation
algorithm affects which joint distributions may be simulated. Even in LD
spaces (e.g., \(d=3\)), there exist valid multivariate distributions
with \emph{feasible} Pearson correlation matrices that NORTA cannot
match exactly \citep{LH75}. This occurs when the bivariate
transformations are applied to find the input correlation matrix, yet
when combined the resultant matrix is indefinite. These situations do
occur, even using exact analytic calculations. Such problematic target
correlation matrices are termed \emph{NORTA defective}.

@@

There is hope, however, as \citet{GH02} also showed that replacing an
indefinite input correlation matrix with a close proxy will give
approximate matching to the target --- with adequate performance for
moderate \(d\). This provides guidance that our nearest positive
definite (PD) augmented approach has promise provide adequate accuracy
if our input matching scheme returns an indefinite matrix.

\hypertarget{rand-vec-gen}{%
\subsection{bigsimr rvec: Random vector generator}\label{rand-vec-gen}}

We now describe our algorithm to generate random vectors, which extends
NORTA to HD via nearest correlation matrix replacement and provides
rank-based dependency matching:

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Mapping step, either

  \begin{itemize}
  \tightlist
  \item
    Convert the target Spearman correlation matrix \(R_S\) to the
    corresponding MVN Pearson correlation \(R_X\). Alternatively,
  \item
    Convert the target Kendall \(\tau\) matrix \(R_K\) to the
    corresponding MVN Pearson correlation \(R_X\). Alternatively,
  \item
    Convert the target Pearson correlation matrix to \(R_P\) to the
    corresponding, approximate MVN Pearson correlation \(R_X\).
  \end{itemize}
\item
  Check admissibility and, if needed, compute the nearest correlation
  matrix.

  \begin{itemize}
  \tightlist
  \item
    Check that \(R_X\) is a correlation matrix, a positive definite
    matrix with 1's along the diagonal.
  \item
    If \(R_X\) is a correlation matrix, the input matrix is
    \emph{admissible} in this scheme. Otherwise,
  \item
    Replace \(R_X\) with the nearest correlation matrix \(\tilde{R}_X\),
    in the Frobenius norm.
  \end{itemize}
\item
  Gaussian copula
